Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-18 _tex/index.tex
Built site for gh-pages

\begin{abstract}
Dichotomous Comparative Judgment (DCJ, \citet{Pollitt_2012a},
\citet{Pollitt_2012b}) requires judges to evaluate the relative
manifestation of traits between pairs of stimuli, resulting in a
dichotomous outcome indicating which stimulus exhibits the trait more
strongly. Research has demonstrated DCJ's effectiveness and reliability
in various domains
\citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
However, the literature lacks a clear and transparent depiction of the
plausible mechanisms underlying DCJ data. Specifically, there is no
detail explanation of how the different assessment factors can
potentially influence the observed DCJ data. This study aims to fill
this gap by applying the framework of causal analysis and Directed
Acyclic Graphs (DAG; \citet{Pearl_2009}). Using this framework, the
study will construct a scientific model to elucidate the causal
assumptions and mechanisms inherent the system. This model will enable
researchers to draw inferences about causal relationships from DCJ data.
Subsequently, the study will translate this model into a probabilistic
statistical model, aiming to derive statistical estimands for different
targets of inference. The outcomes of this study will inform the
planning of DCJ experiments and hold significance for researchers or
analysts involved in education and assessment procedures who implement
the DCJ methodology.
\end{abstract}

@@

In contemporary contexts, Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} primarily refers to the method of
\emph{Dichotomous} Comparative Judgment
\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
the relative manifestation of a \emph{trait} within a pair of stimuli.
This assessment results in a dichotomous value indicating which stimulus
possesses a higher degree of the trait. After different judges perform
multiple rounds of pairwise comparisons, an outcome vector is produced.
This vector is modeled using the Bradley-Terry-Luce model
\citep[BTL,][]{Bradley_et_al_1952, Luce_1959, Pollitt_2012a}, which
produces a latent variable corresponding to the trait of interest. This
latent variable is then used to rank the stimuli from lowest to highest
or to evaluate the influence of certain variables on the stimuli's
positions in the ranking.

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by the works
of \citet{Pollitt_2012b}, \citet{Bartholomew_et_al_2018},
\citet{vanDaal_et_al_2016}, \citet{Lesterhuis_2018}, and
\citet{Bartholomew_et_al_2020}. However, its application transcends
education, as exemplified by the work of \citet{Boonen_et_al_2020}. The
methodology has also evolved to include multiple, as opposed to pairwise
comparisons \citep{Luce_1959, Placket_1975}, and to accommodate
comparisons with ordinal outcomes \citep{Tutz_1986, Agresti_1992}.
Overall, research suggests that DCJ offers an alternative and efficient
approach to measurement and evaluation, characterized by its reliability
and validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature lacks
a transparent depiction of the DCJ system and the plausible mechanisms
that give rise to DCJ data. Particularly, there is no detailed
explanation of how different assessment factors can potentially
influence the observed DCJ data.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and contribute to the reliability of
the DCJ method. These factors include the number and characteristics of
the stimuli, their \emph{proximity} in terms of the assessed trait, the
number of comparison per stimulus, and the pairing algorithm used.
Furthermore, since the method relies on judges' assessments, the number
and characteristics of judges, their \emph{discrimination} abilities,
and the number of comparisons per judge also play pivotal roles.
Moreover, when the stimuli represent sub-units of higher-levels units,
factors such as the number and characteristics of these units, along
with their \emph{proximity} in terms of the assessed trait, can
significantly influence the outcome. An example of this can be found in
\citet{vanDaal_et_al_2016}, where the authors assessed the \emph{skills
in academic writing} (trait) of Flemish university students, utilizing
multiple argumentative essays (stimuli, sub-units) originating from
various students (units).

>>>>

\begin{abstract}
Dichotomous Comparative Judgment
\citep[DCJ,][@Pollitt\_2012b]{Pollitt_2012a} requires judges to evaluate
the relative manifestation of traits between pairs of stimuli, resulting
in a dichotomous outcome indicating which stimulus exhibits the trait
more strongly. Research has demonstrated DCJ's effectiveness and
reliability in various domains
\citep{Pollitt_2012b, Bartholomew_et_al_2018, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2020, Boonen_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature lacks
a transparent depiction of the DCJ system and the plausible mechanisms
that generate the DCJ data. Particularly, there is no detailed
explanation of how different assessment factors can potentially
influence the observed DCJ data. This study aims to fill this gap by
applying the framework of causal analysis and Directed Acyclic Graphs
{[}DAG; \citet{Pearl_2009}{]}. Using this framework, the study will
construct a scientific model to elucidate the causal assumptions and
mechanisms inherent the system. This model will enable researchers to
draw inferences about causal relationships from DCJ data. Subsequently,
the study will translate this model into a probabilistic statistical
model, aiming to derive statistical estimands for different targets of
inference. The outcomes of this study will inform the planning of DCJ
experiments and hold significance for researchers or analysts involved
in education and assessment procedures who implement the DCJ
methodology.
\end{abstract}

@@

In contemporary contexts, Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} primarily refers to the method of
\emph{Dichotomous} Comparative Judgment
\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
the relative manifestation of a \emph{trait} within a pair of stimuli.
This assessment results in a dichotomous value indicating which stimulus
possesses a higher degree of the trait. After different judges perform
multiple rounds of pairwise comparisons, an outcome vector is produced.
This vector is modeled using the Bradley-Terry-Luce model
\citep[BTL,][]{Bradley_et_al_1952, Luce_1959}, which creates a score
that corresponds with the trait of interest. This score is then used to
rank the stimuli from lowest to highest or to evaluate the influence of
certain variables on the stimuli's positions in the ranking.

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, there is no clear
representation in the literature of the plausible mechanisms that
generate DCJ data. Particularly, there is no depiction of the complexity
and the underlying assumptions of the DCJ system, nor how different
assessment factors can potentially influence the observed DCJ outcome.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the DCJ outcome. These
factors include the number and characteristics of the stimuli, their
\emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
university students' skills in academic writing, utilizing multiple
argumentative essays (stimuli, sub-units) originating from various
students (units).

Although several studies have examined the individual impact of these
factors on the method's reliability, including \citet{Bramley_2015},
\citet{Pollitt_2012b}, \citet{Bramley_et_al_2019},
\citet{Verhavert_et_al_2019}, \citet{Crompvoets_et_al_2022},
\citet{vanDaal_et_al_2017}, and \citet{Gijsen_et_al_2021}, to the best
of the authors' knowledge, none have provided such transparent depiction
of DCJ system and the plausible mechanisms that generate the DCJ
outcome. This study aims to fill this gap by utilizing the framework of
causal analysis and Directed Acyclic Graphs {[}DAG; \citet{Pearl_2009};
\citet{Pearl_et_al_2016}{]}. Using this framework, the study will
construct a scientific model to elucidate the underlying assumptions of
the DCJ system, providing plausible mechanisms of how the DCJ data is
generated. This model will enable researchers to draw inferences about
plausible causal relationships within the DCJ system. Furthermore, using
a minimal set of assumptions from the framework, the study will
translate the scientific model into a probabilistic statistical model,
aiming to derive statistical estimands for different targets of
inference. Ultimately, the results of this study could inform the
planning of DCJ experiments and hold significance for researchers or
analysts involved in education and assessment procedures who implement
the DCJ methodology.
